FraCaS: Temporal Analysis

### 2020-10-01 paper2020/paper.tex
deal with review 1

\section{Background}
\label{sec:background}

\paragraph{GF}
The grammatical framework (GF, \citet{ranta_grammatical_2004}) is a
powerful parser generator for natural languages, based on
type-theoretical abstract grammars.  The present work leverages a
syntactic representation of the FraCaS test suite in GF abstract
syntax~\citep{Ljunglof:2012}. Thanks to this, we skip the parsing
phase and avoid any syntactic ambiguity.

@@

\paragraph{Coq}
Coq is an interactive theorem prover (proof assistant). It is based on
the calculus of co-inductive constructions~\citep{werner_une_1994} We
do not use any co-induction (or even induction) in this paper, relying
on the pure lambda-calculus inner core of Coq.  Coq is a very powerful
reasoning engine that makes it fit for implementing natural language
semantics.  Coq supports, as already mentioned, dependent typing, but
also subtyping. Both concepts are instrumental in expressing NL
semantics\citet{chatzikyriakidis_natural_2014}.  Besides, on a more
practical side, it works well for the the task of NLI, when the latter
is formalized as a theorem proving task: its many tactics mean that
many tasks in theorem proving are trivialized. In particular, all
problems of time-intervals inclusion, which occur in every temporal
problems, are solved with Coq's linear arithmetic tactic.

\paragraph{Montagovian Semantics}

Our interpretation is cast within the framework of a largely standard
montagovian semantics. In brief, sentences are interpreted as
propositions, verbs and noun-phrases as predicates.  We use
type-raising of noun-phrases, to support quantifiers~\citep{montague_proper_1974}.

We support additionally the basic constructions and phenomena present
in the testsuite, including adjectives, adverbs, nouns, verbs,
anaphora, etc. We direct the reader to our previous work for details
\ifanon (citations anonymized) \else
\citet{bernardy_type-theoretical_2017} \fi, but the particular
treatment of such phenomena is essentially independent from our
treatment of time:
in this paper we simply ignore these aspects beyond the fact that they
are handled correctly in most of the cases of the FraCaS testsuite.

\paragraph{Dynamic Semantics}
We additionally assume support for (non-temporal) anaphora, using a
monadic-based dynamic semantics, generally following the state of the
art in this matter
\citep{unger_dynamic_2011,charlow_monadic_2015,charlow_modular_2017}.
Our particular implementation does not handle every case \ifanon
(citations anonymized) \else \citet{bernardy_wide-coverage_2019} \fi
but non-temporal anaphoroi in the testsuite are generally resolved as
they should be: on the whole accuracy is not affected significantly by
issues in this subsystem.

@@

\paragraph{Temporal Semantics}

In this paper we borrow several notions from classical temporal
semantics such as ``stative'', ``achievement'', ``activity'', etc.,
even though our definitions do not perfectly match the classical
ones. We explain our precise meaning for these terms in the body of the
paper. Nevertheless, we refer the reader to \citet{steedman:2000} for
an extensive review of formal temporal semantics.

@@

We adjust the montagovian semantics so that the interpretation of
every category (propositions, verb phrases, etc.) takes a
\emph{temporal context} as an additional parameter, which serves as a time
reference for the interpretation of all time-dependent semantics
within the phrase.  This context propagates through the compositional
interpretation down to lexical items with atomic representation
(verbs). By default, every interpretation passes the temporal context
down to its components without changing it. However some key elements
will act on it on nontrivial ways, which we proceed to detail below.

@@

\item[existentially quantifying] The majority of temporal adverbs
  existentially quantify over a timespan. Examples include ``since
  1991'', ``in 1996'', ``for two years'', etc. The common theme is to
  introduce the interval and then restrict its bounds or its duration
  in some way. Sometimes the restriction is an equality, as in ``for
  exactly two hours''.

@@

\paragraph{(Un)repeatable Achievements}
\label{sec:repeatable-achievements}
The principle of using unicity of action interacts well with the usual
interpretation of quantifiers (and anaphora).  Indeed, using it, we
can refute problem (279), as expected by the testsuite:

@@

Finally we note that not all verbs are subject to the
stative/achievement distinction induced by the progressive. For
example, the phrases ``John ran'' and ``John is running'' appear to be
logically equivalent, for entailment purposes.

@@

Our dynamic and temporal semantics gives the following interpretation
for \verb!s_279_1_p! implies \verb!s_279_3_h!.
{\small
\begin{verbatim}
FORALL (fun a=>novel_N a)
(fun a=>(exists (b: Time),
((exists (c: Time),
(IS_INTERVAL Date_19910101 b /\
 IS_INTERVAL c Date_19911231 /\
 IS_INTERVAL b c /\
 appTime b c (write_V2 a)
   (PN2object smith_PN))))) ->
Not (exists (f: Time),
((exists (g: Time),
 (IS_INTERVAL Date_19920101 f /\
  IS_INTERVAL g Date_19921231 /\
  IS_INTERVAL f g /\
  appTime f g (write_V2 a) 
    (PN2object smith_PN)))))).
\end{verbatim}
}
In the above, one should remark the top-level quantification over the
novel (as explained in \cref{sec:our-semantics}), the quantification
over time intervals as individual timepoints, and the use of custom
operators for several constructions (\verb!FORALL!, \verb!Not!, \verb!IS_INTERVAL!,
\verb!appTime!). This use of custom operators is useful for several
generalizations (for example, we have quantifiers such as \verb!MOST! in
addition to \verb!FORALL!.)

>>>>

\paragraph{Temporal Semantics}
\todo{Fold this in some other section}
In this paper we borrow several notions from classical temporal
semantics such as ``stative'', ``achievement'', ``activity'', etc.,
even though our definitions do not perfectly match the classical
ones. We explain our precise meaning for these terms in the body of the
paper. Nevertheless, we refer the reader to \citet{steedman:2000} for
an extensive review of formal temporal semantics.

\section{Temporal-Semantics in a Logic-based NLI System}
\label{sec:background}

Our temporal analysis places itself in the context of a complete NLI
system -- which is why we can test it on the FraCaS suite. In this
section we give a brief overview of the phases of the system,
referring the reader to published work for details.

\paragraph{GF}
The first phase of the system, parsing, is taken care of by the 
Grammatical Framework (GF, \citet{ranta_grammatical_2004}), which is a
powerful parser generator for natural languages, based on
type-theoretical abstract grammars.  The present work leverages a
syntactic representation of the FraCaS test suite in GF abstract
syntax~\citep{Ljunglof:2012}. Thanks to this, we skip the parsing
phase and avoid any syntactic ambiguity.

@@

\paragraph{Dynamic Semantics}
Parse trees are then processed by a dynamic semantic component.
Its role is essentially to support  (non-temporal) anaphora, using a
monadic-based dynamic semantics, generally following the state of the
art in this matter
\citep{unger_dynamic_2011,charlow_monadic_2015,charlow_modular_2017}.
Our particular implementation does not handle every case \ifanon
(citations anonymized) \else \citet{bernardy_computational_2020,bernardy_wide-coverage_2019} \fi
but non-temporal anaphoroi in the testsuite are generally resolved as
they should be: on the whole accuracy is not affected significantly by
issues in this subsystem.

@@

\paragraph{Montagovian Semantics}
Non-withstanding special support for anaphora, the core of the translation of syntax trees
to logical form follows a largely standard
montagovian semantics. In brief, sentences are interpreted as
propositions, verbs and noun-phrases as predicates.  We use
type-raising of noun-phrases, to support
quantifiers~\citep{montague_proper_1974}.

We support additionally the basic constructions and phenomena present
in the testsuite, including adjectives, adverbs, nouns, verbs,
anaphora, etc. The method is outlined by
\citet{montague_english_1970,montague_proper_1973}, but we direct the
reader to our previous work for details \ifanon (citations anonymized)
\else \citet{bernardy_type-theoretical_2017,bernardy_wide-coverage_2019} \fi, but the particular
treatment of such phenomena is essentially independent from our
treatment of time: in this paper we simply ignore these aspects beyond
the fact that they are handled correctly in most of the cases of the
FraCaS testsuite.

\paragraph{Inference using Coq}
Logical forms are then fed to the Coq
Coq interactive theorem prover (proof assistant). Coq is based on
the calculus of co-inductive constructions~\citep{werner_une_1994} We
do not use any co-induction (or even induction) in this paper, relying
on the pure lambda-calculus inner core of Coq.  Coq is a very powerful
reasoning engine that makes it fit for implementing natural language
semantics.  Coq supports, as already mentioned, dependent typing, but
also subtyping. Both concepts are instrumental in expressing NL
semantics\citet{chatzikyriakidis_natural_2014}.  Besides, on a more
practical side, it works well for the the task of NLI, when the latter
is formalized as a theorem proving task: its many tactics mean that
many tasks in theorem proving are trivialized. In particular, all
problems of time-intervals inclusion, which occur in every temporal
problems, are solved with Coq's linear arithmetic tactic.

@@

We adjust the montagovian semantics so that the interpretation of
every category (propositions, verb phrases, etc.) takes a
\emph{temporal context} as an additional parameter, which serves as a time
reference for the interpretation of all time-dependent semantics
within the phrase. (While some categories do not need this temporal context, we pass it everywhere for consistency.)  This context propagates through the compositional
interpretation down to lexical items with atomic representation
(verbs). By default, every interpretation passes the temporal context
down to its components without changing it. However some key elements
will act on it on nontrivial ways, which we proceed to detail below.

@@

\item[existentially quantifying] The majority of temporal adverbs
  existentially quantify over a timespan. Examples include ``since
  1991'', ``in 1996'', ``for two years'', etc. The common theme is to
  introduce the interval and then restrict its bounds or its duration
  in some way. Sometimes the restriction is an equality, as in ``for
  exactly two hours''. In the following example we show the inclusion
  constraint, for ``in 1992''.

@@

\paragraph{(Un)repeatable Achievements}
\label{sec:repeatable-achievements}
The principle of using unicity of action interacts well with the usual
interpretation of existential quantifiers (and anaphora).  Indeed, using it, we
can refute problem (279), as expected by the testsuite:

@@

Finally we note that not all verbs are subject to the
stative/achievement distinction induced by the progressive. For
example, the phrases ``John ran'' and ``John was running'' appear to be
logically equivalent, for entailment purposes.

@@

Our dynamic and temporal semantics gives the following interpretation
for \verb!s_279_1_p! implies \verb!s_279_3_h!.
{\small
\begin{verbatim}
FORALL (fun a=>novel_N a)
(fun a=>(exists (b: Time),
((exists (c: Time),
(IS_INTERVAL Date_19910101 b /\
 IS_INTERVAL c Date_19911231 /\
 IS_INTERVAL b c /\
 appTime b c (write_V2 a)
   (PN2object smith_PN))))) ->
Not (exists (f: Time),
((exists (g: Time),
 (IS_INTERVAL Date_19920101 f /\
  IS_INTERVAL g Date_19921231 /\
  IS_INTERVAL f g /\
  appTime f g (write_V2 a) 
    (PN2object smith_PN)))))).
\end{verbatim}
}
In the above, one should remark the top-level quantification over the
novel (as explained in \cref{sec:our-semantics}), the quantification
over time intervals as individual timepoints, and the use of custom
operators for several constructions (\verb!FORALL!, \verb!Not!, \verb!IS_INTERVAL!,
\verb!appTime!). This use of custom operators is useful for several
generalizations (for example, we have quantifiers such as \verb!MOST! in
addition to \verb!FORALL! --- see \ifanon (citation anonymized)
\else \citet{bernardy_type-theoretical_2017} \fi)
